---
title: "IPW4CausalInference"
author: "Awol Seid"
date: "2025-02-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Point treatment example

A point treatment data with measurements made in 1000 individuals on a continuous confounder $L$, a dichotomous exposure $A$ and a continuous outcome $Y$ , using

  - $L\sim N(10,5)$
  - $\log\left(\frac{P[A=1]}{1-P[A=1]}\right)=-10+L$
  - $Y = 10A+0.5L+N(-10,5)$

The true parameter for the marginal causal effect of $A$ on $Y$ is 10. 
```{r}
library(ipw)

set.seed(16)

n <- 1000
n
simdat <- data.frame(l = rnorm(n, 10, 5))
simdat[1:10,]
```
```{r}
a.lin <- simdat$l - 10
a.lin[1:10]
```
```{r}
pa <- exp(a.lin)/(1 + exp(a.lin))
pa[1:10]
```
```{r}
simdat$a <- rbinom(n, 1, prob = pa)
simdat$a[1:10]
```



```{r}
simdat$y <- 10*simdat$a + 0.5*simdat$l + rnorm(n, -10, 5)
simdat[1:10,]
```
```{r}
summary(lm(simdat$y~simdat$a+simdat$l))
```
We can estimate inverse probability weights to correct for the confounding.

$$w_i=\frac{P(A_i=a_i)}{P(A_i=a_i|L_i=l_i)}$$
```{r}
model_A = glm(a~1, family="binomial", data=simdat)
summary(model_A)
pred_prob_A = model_A$fitted.values
pred_prob_A[1:10]
prob_A <- ifelse(simdat$a == 1, pred_prob_A, 1 - pred_prob_A)
prob_A[1:10]
```

```{r}
model_AgL = glm(a~l, family="binomial", data=simdat)
summary(model_AgL)
pred_prob_AgL = model_AgL$fitted.values
pred_prob_AgL[1:10]
prob_AgL <- ifelse(simdat$a == 1, pred_prob_AgL, 1 - pred_prob_AgL)
prob_AgL[1:10]
```
```{r}
wi = prob_A/prob_AgL
summary(wi)
```
Weighting the original observations by the stabilized weights $wi$ to adjust for confounding,
we can fit the MSM, estimating the marginal causal effect of $A$ on $Y$ ,
$$Y=\alpha+\beta A$$

```{r}
library(survey)
msm <- (svyglm(y ~ a, design = svydesign(~ 1, weights = ~ wi, data = simdat)))
summary(msm)
confint(msm)
```